---
title: "Test of Biomass as Covariate in Aster Models for Lifetime Fitness"
author: "Mason W. Kulbaba & Ruth G. Shaw"
date: "Last Updated June 28, 2020"
output: pdf_document
---

# Introduction

The following code tests, on the advice of Dr. Daniel Bolnick and Dr. Lynda Delph, the effect of biomass on female and male fitness, as determined with aster models.
        
## Preliminaries   

Set working directory and load data.
```{r global options, echo=FALSE}
#setwd()
```

                                                                                           
Examine data
```{r }
fin<- read.csv("data/aster.dat.csv")

names(fin)

head(fin)
```

Make sure Den (density treatment), Gen (N~e~ treatment), plotID, plantID, and familyID are all classified as factors. Otherwise
```{r}
fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)
```


## Setting up for aster analyses

Load aster package (make sure you have most current version)

```{r, message = FALSE, warning = FALSE}
library(aster)
```

Begin by naming variables that will be used in the graphical model of the aster analyses:

**flw** - total number of flowers produced
**frt** - total number of fruits produced
**frt.2** - subsetted number of fruits collected
**seeds** - total number of seeds collected from subsetted fruits

```{r}
vars<- c( "flw", "frt", "frt.2","seeds")
```

Reshape the data so that all response variables are located in a single vector, in a new data
set called "redata"
```{r}
redata <- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
```

Designate the terminal fitness variable "seeds" (make it numberic), and then add it to the reshaped data
```{r}
fit <- grepl("seeds", as.character(redata$varb))
fit<- as.numeric(fit)
redata$fit <- fit
```

Check
```{r}
with(redata, sort(unique(as.character(varb)[fit == 0])))
with(redata, sort(unique(as.character(varb)[fit == 1])))
```

Add a variable "root" to redata, where value is 1. This is the "starting point" of the aster graphical model (i.e. a seed planted)
```{r}
redata<- data.frame(redata, root=1)
```

Set up the graphical model and designate the statistical distribution for each node.
This graphical model has four nodes (in object `pred`) described earlier. Statistical family for each node is described by object `fam`.
```{r}
pred<- c(0,1,2,3)
fam<- c(1,2,2,2)
```

Show distribution family for each node
```{r}
sapply(fam.default(), as.character)[fam]
```



# Main Aster Analyses for Female Fitness (Seeds set)

First aster analysis with only fitness data. Note, `aster` reads the `redata` version of the data.

```{r}
aout<- aster(resp~varb, pred, fam, varb, id, root, data=redata)

aout1<- aster(resp~varb + fit:(mass.a), pred, fam, varb, id, root, data=redata)

aout2<- aster(resp~varb + fit:(mass.a + familyID), pred, fam, varb, id, root, data=redata)

aout3<- aster(resp~varb + fit:(mass.a + familyID + Gen), pred, fam, varb, id, root, data=redata)

aout4<- aster(resp~varb  + fit:(mass.a + familyID + Gen + Den), pred, fam, varb, id, root, data=redata)

aout5<- aster(resp~varb + fit:(mass.a + familyID + Gen + Den + Gen*Den) , pred, fam, varb, id, root, data=redata)


anova(aout, aout1, aout2, aout3, aout4, aout5)
```

### Interpretation
Sequential addition of `familyID`, `Gen` (Ne), `Den` (density), `Gen * Den` (Gen by Den interaction), and finally `mass.a` (above ground biomass) consistently explained more variation in total lifetime fitness. Next, move on to specific models to estimate fitness across treatments.


## Test of effectos of biomass on male fitness (relative number of seeds sired)

Load data with relative number of seeds set (`rel.seeds`) and sired (`sires`).
```{r}
fin<- read.csv("C:/Users/Mason Kulbaba/Dropbox/git/density-gen-Ne/data/aster.sire.dat.csv")

```

Perform preliminary tasks of factor class setting, and organizing of data for density-treatment specific analyses, as the number of seeds sired was only estimated for the high N~e~ treatment.
```{r}
fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)


#subset data to density-specific treatments and 'drop' unused levels
datH<- subset(fin, Den=="H")
datH$familyID<- droplevels(datH$familyID)
datM<- subset(fin, Den=="M")
datM$familyID<- droplevels(datM$familyID)
datL<- subset(fin, Den=="L")
datL$familyID<- droplevels(datL$familyID)

#set vars
vars<- c( "flw", "frt", "sires")

redata<- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")


```

Designate fitness var
```{r}
#Designation of fitness variable
fit <- grepl("sires", as.character(redata$varb))
fit<- as.numeric(fit)

#add fit to each of three redata files (one for each ensity treat)
redata$fit <- fit

```

Add a variable "root" to redata, where value is 2

```{r}
redata<- data.frame(redata, root=2)
```


Set graphical model and statistical family for each node
```{r}
pred<- c(0,1,2)
fam<- c(1,2,2)

sapply(fam.default(), as.character)[fam]
```

Build aster models
```{r}
aout.sire<- aster(resp~varb, pred, fam, varb, id, root, data=redata)

aout.sire2<- aster(resp~varb + fit:(mass.a), pred, fam, varb, id, root, data=redata)

aout.sire3<- aster(resp~varb + fit:(mass.a + familyID), pred, fam, varb, id, root, data=redata)

aout.sire4<- aster(resp~varb + fit:(mass.a + familyID + Den), pred, fam, varb, id, root, data=redata)

anova(aout.sire, aout.sire2, aout.sire3, aout.sire4)
```






