---
title: "Lifetime fitness through female and male function: the influence of density and genetically effective population size"
author: "Mason W. Kulbaba & Ruth G. Shaw"
date: "April 5, 2019"
output: pdf_document
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
```

## Introduction

The following code performs fixed-effects aster analyses on data examining the effects
of density and effective genetic population size (Ne) on female (seeds seet) fitness. 
LM analysis of biomass (above and below ground) and number of aborted ovules follows  
the aster analyses.   

Please send any questions to Mason Kulbaba (mason.kulbaba@gmail.com)
        
## Preliminaries   

Set working directory and load data.
```{r global options, echo=FALSE}

```

                                                                                           
Examine data
```{r }
fin<- read.csv("data/aster.dat.csv")

names(fin)

head(fin)
```

Make sure Den (density treatment), Gen (Ne treatment), plotID, plantID, and familyID are all classified as factors. Otherwise
```{r}
fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)
```


## Setting up for aster analyses

Load aster package (make sure you have most current version)

```{r}
library(aster)
```

Begin by naming variables that will be used in the graphical model of the aster analyses:

flw - total number of flowers produced
frt - total number of fruits produced
frt.2 - subsetted number of fruits collected
seeds - total number of seeds collected from subsetted fruits

```{r}
vars<- c( "flw", "frt", "frt.2","seeds")
```

Reshape the data so that all response variables are located in a single vector, in a new data
set called "redata"
```{r}
redata <- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
```

Designate the terminal fitness variable "seeds" (make it numberic), and then add it to the reshaped data
```{r}
fit <- grepl("seeds", as.character(redata$varb))
fit<- as.numeric(fit)
redata$fit <- fit
```

Check
```{r}
with(redata, sort(unique(as.character(varb)[fit == 0])))
with(redata, sort(unique(as.character(varb)[fit == 1])))
```

Add a variable "root" to redata, where value is 1. This is the "starting point" of the aster graphical model (i.e. a seed planted)
```{r}
redata<- data.frame(redata, root=1)
```

Set up the graphical model and designate the statistical distribution for each node.
This graphical model has four nodes (in object `pred`) described earlier. Statistical family for each node is described by object `fam`.
```{r}
pred<- c(0,1,2,3)
fam<- c(1,2,2,2)
```

Show distribution family for each node
```{r}
sapply(fam.default(), as.character)[fam]
```

## Aster Analyses

First aster analysis with only fitness data. Note, `aster` reads the `redata` version of the data.

```{r}
aout<- aster(resp~varb , pred, fam, varb, id, root, data=redata)

summary(aout, show.graph=T)
```

Add density to the model as a fixed effect. Note that we follow the "no naked predictors" rule with `fit:(Den)`
```{r}
aout.d<- aster(resp~varb + fit:(Den), pred, fam, varb, id, root, data=redata)

summary(aout.d, show.graph=T)
```

Do not try to determine the significance `Density` from the `summary` statement. These are not reliable. Instead, perform a liklihood ratio test. The results of the liklihood ratio test show that the model with `Density` explains significantly more variation than the model without.

```{r}
anova(aout, aout.d)
```


#### Make Up Data

Simulate regression data, and do the regression.
```{r}
n <- 50
x <- seq(1, n)
a.true <- 3
b.true <- 1.5
y.true <- a.true + b.true * x
s.true <- 17.3
y <- y.true + s.true * rnorm(n)
out1 <- lm(y ~ x)
summary(out1)
```

#### Figure with Code to Make It Shown

The following figure is produced by the following code
```{r fig.align='center', fig.cap='Simple Linear Regression', label='regression'}

mydata <- data.frame(x, y)
plot(mydata)
``````