---
title: "Lifetime fitness through female and male function: the influence of density and genetically effective population size"
author: "Mason W. Kulbaba & Ruth G. Shaw"
date: "Last Updated April 11, 2019"
output: pdf_document
toc: true
toc_depth: 3
---

# Introduction

The following code performs fixed-effects aster analyses on data examining the effects
of density and effective genetic population size (N~e~) on female (seeds seet) fitness. 
LM analysis of biomass (above and below ground) and number of aborted ovules follows  
the aster analyses.   

Please send any questions to Mason Kulbaba (mason.kulbaba@gmail.com)
        
## Preliminaries   



Set working directory and load data.
```{r global options, echo=FALSE}

```

                                                                                           
Examine data
```{r }
fin<- read.csv("data/aster.dat.csv")

names(fin)

head(fin)
```

Make sure Den (density treatment), Gen (N~e~ treatment), plotID, plantID, and familyID are all classified as factors. Otherwise
```{r}
fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)
```


## Setting up for aster analyses

Load aster package (make sure you have most current version)

```{r}
library(aster)
```

Begin by naming variables that will be used in the graphical model of the aster analyses:

flw - total number of flowers produced
frt - total number of fruits produced
frt.2 - subsetted number of fruits collected
seeds - total number of seeds collected from subsetted fruits

```{r}
vars<- c( "flw", "frt", "frt.2","seeds")
```

Reshape the data so that all response variables are located in a single vector, in a new data
set called "redata"
```{r}
redata <- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
```

Designate the terminal fitness variable "seeds" (make it numberic), and then add it to the reshaped data
```{r}
fit <- grepl("seeds", as.character(redata$varb))
fit<- as.numeric(fit)
redata$fit <- fit
```

Check
```{r}
with(redata, sort(unique(as.character(varb)[fit == 0])))
with(redata, sort(unique(as.character(varb)[fit == 1])))
```

Add a variable "root" to redata, where value is 1. This is the "starting point" of the aster graphical model (i.e. a seed planted)
```{r}
redata<- data.frame(redata, root=1)
```

Set up the graphical model and designate the statistical distribution for each node.
This graphical model has four nodes (in object `pred`) described earlier. Statistical family for each node is described by object `fam`.
```{r}
pred<- c(0,1,2,3)
fam<- c(1,2,2,2)
```

Show distribution family for each node
```{r}
sapply(fam.default(), as.character)[fam]
```

# Main Aster Analyses for Female Fitness (Seeds set)

First aster analysis with only fitness data. Note, `aster` reads the `redata` version of the data.

```{r}
aout<- aster(resp~varb , pred, fam, varb, id, root, data=redata)

summary(aout, show.graph=T)
```

Add density to the model as a fixed effect. Note that we follow the "no naked predictors" rule with `fit:(Den)`
```{r}
aout.d<- aster(resp~varb + fit:(Den), pred, fam, varb, id, root, data=redata)

summary(aout.d, show.graph=T)
```


Do not try to determine the significance `Density` from the `summary` statement. These are not reliable. Instead, perform a liklihood ratio test. The results of the liklihood ratio test show that the model with `Density` explains significantly more variation than the model without.

```{r}
anova(aout, aout.d)
```

Now, add density and N~e~ (`Gen`) to the model, and perform a liklihood ratio test
```{r}
aout.dg<- aster(resp~varb + fit:(Den + Gen), pred, fam, varb, id, root, data=redata)

summary(aout.dg, show.graph=T)

anova(aout.d, aout.dg)
```


Include interaction between `Den` & `Gen` and test for significance of interaction with liklihood ratio test.
```{r}
aout.dg2<- aster(resp~varb + fit:(Den + Gen + Den*Gen), pred, fam, varb, id, root, data=redata)

summary(aout.dg2)

anova(aout.dg, aout.dg2)
```

Finally, include plotID to model and perform yet anohter liklihood ratio test
```{r}
aoutc<- aster(resp~varb + fit:(plotID + Den + Gen + Den*Gen), pred, fam, varb, id, root, data=redata)

summary(aoutc, show.graph=T)


anova(aout.dg2, aoutc)
```

That was fun. The results of these aster models (of entire data set) and liklihood ratio tests are presetned in Table 1 of the manuscript. The next steps will produce mean fitness and standard errors for these factors.

## Calculation of Mean Fitness and Standard Errors

As the effects of `GEN` was significant in the above analyses, we can divide the data into high and low N~e~ data sets, and perform additional analsyis to calcualte mean fitness and standard errors for the density treatmetns. The aster analyses for high and low N~e~ are performed in parallel below (i.e. each step is performed twice, once for high N~e~ and once for low N~e~ analses).

First, isloate high (HG) and low (LG) data from the main redata file, and drop unused levels. Therefore, don't have to do "reshape" data step. `HG`/`hg` = High N~e~, `LG`/`lg` = Low N~e~
```{r}
redataHG<- subset(redata, Gen=="HG")
redataLG<- subset(redata, Gen=="LG")

redataHG<- droplevels(redataHG)
redataLG<- droplevels(redataLG)
```

Perform aster analysis on `HG` and `LG` data with just fitness data (no predictors), then add `Den` and perform a liklihood ratio test.
```{r}
aoutHG<- aster(resp~varb, pred, fam, varb, id, root, data=redataHG)
aoutHG2<- aster(resp~varb + fit:(Den), pred, fam, varb, id, root, data=redataHG)

summary(aoutHG, show.graph = T)
summary(aoutHG2, show.graph=T)

anova(aoutHG, aoutHG2)
```


Perform same sequence of analyses on `LG` dataset 
```{r}
aoutLG<- aster(resp~varb, pred, fam, varb, id, root, data=redataLG)
aoutLG2<- aster(resp~varb + fit:(Den), pred, fam, varb, id, root, data=redataLG)

summary(aoutLG, show.graph = T)
summary(aoutLG2, show.graph=T)

anova(aoutLG, aoutLG2)

```

The effectds of density (`Den`) was significant in both the high (`HG`) and low (`LG`) effective genetic population size data sets. 

First step is to generate MLE of saturated model mean value parameter vector: mu. Again, all step are performed twice, once for high and once for low N~e~ data. Because we want treatment-level estiamtes of fitness, we generate these estimates from the analyses that included density: `aoutHG` and `aoutLG`
```{r}
pout.HG<- predict(aoutHG, se.fit=TRUE)

pout.LG<- predict(aoutLG, se.fit=TRUE)
```

Make design matrix data.frame of indivudals for each density level (low, med., high), that has a 1 for each element of the matrix. These will eventually be replaced with actual fitness values in later steps.
```{r}
fred.hg <- data.frame( Den=levels(redataHG$Den), flw=1, frt=1,frt.2=1, seeds=1,root = 1)

fred.lg <- data.frame( Den=levels(redataLG$Den), flw=1, frt=1,frt.2=1, seeds=1,root = 1)
```

Reshape the design matrix just as the actual data
```{r}
renewdata.hg <- reshape(fred.hg, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")

renewdata.lg <- reshape(fred.lg, varying = list(vars),
                        direction = "long", timevar = "varb",
                        times = as.factor(vars), v.names = "resp")
```


Make character string from "varb" of renewdata without actual values (i.e., the layers of varb in renewdata), and add it to each renewdata object
```{r}
layer<- gsub("[0-9]", "", as.character(renewdata.hg$varb))


renewdata.hg<- data.frame(renewdata.hg, layer= layer)
renewdata.lg<- data.frame(renewdata.lg, layer= layer)
```



Add "seeds" in new layer column of renewdata as numeric, called fit
Note: only need one `fit` object as it is the same for both High and Low N~e~ data, and add to each renew data file
```{r}
fit<- as.numeric(layer=="seeds")

renewdata.gh<- data.frame(renewdata.hg, fit = fit)
renewdata.lg<- data.frame(renewdata.lg, fit = fit)
```


Rerun prediction of aster analyses, with the reshaped design matrices 
```{r}
pout.hg<- predict(aoutHG2, newdata= renewdata.hg, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)

pout.lg<- predict(aoutLG2, newdata= renewdata.lg, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)
```

Check class of each column in prediction outputs
```{r}
sapply(pout.hg, class)

sapply(pout.lg, class)
```

Lengths of fit and se.fit (12) match row number of renewdata (as should be with predict.aster)
```{r}
sapply(pout.hg, length)
sapply(pout.lg, length)
```

Therefore, we can make 12 CIs, one for each of 4 nodes of graphical model, and 3 density treatmetns (4 nodes x 3 treatments =12 estimates).


Put the parameter estimates into a matrix with individuals in rows and nodes in columns

Extract HG resutls, and produce a 3 x 4 matrix (3 density treatments by 4 nodes)
```{r}
nnode<- length(vars)
sally.hg<- matrix(pout.hg$fit, ncol = nnode)
dim(sally.hg)
```

Name the rows (by `Den` treatments) and columns (as nodes), and view the matrix
```{r}
rownames(sally.hg)<- unique(as.character(renewdata.hg$Den))
colnames(sally.hg)<- unique(as.character(renewdata.hg$varb))

round(sally.hg, 3)
```

Now generate matrix of standard errors, and name rows and columns just as fitness estimates
```{r}
nnode2<- length(vars)
sally2<- matrix(pout.hg$se.fit, ncol = nnode)
dim(sally2)

rownames(sally2)<- unique(as.character(renewdata.hg$Den))
colnames(sally2)<- unique(as.character(renewdata.hg$varb))

round(sally2, 3)
```

Combine estimates with standard errors for only final node: seeds
```{r}
ests<- sally.hg[,grepl("seeds", colnames(sally.hg))]
se<-  sally2[,grepl("seeds", colnames(sally2))]

HG<- cbind(ests, se)

```


Perform the same steps for `LG` results
```{r}
nnode<- length(vars)
sally.lg<- matrix(pout.lg$fit, ncol = nnode)
dim(sally.lg)

rownames(sally.lg)<- unique(as.character(renewdata.lg$Den))
colnames(sally.lg)<- unique(as.character(renewdata.lg$varb))

round(sally.lg, 3)
```

Extract standard errors
```{r}
nnode2<- length(vars)
sally.lg2<- matrix(pout.lg$se.fit, ncol = nnode)
dim(sally.lg2)

rownames(sally.lg2)<- unique(as.character(renewdata.lg$Den))
colnames(sally.lg2)<- unique(as.character(renewdata.lg$varb))

round(sally.lg2, 3)
```

Combine estimates with standard errors for only final node: seeds
```{r}
ests<- sally.lg[,grepl("seeds", colnames(sally.lg))]
se<-  sally.lg2[,grepl("seeds", colnames(sally.lg2))]

LG<- cbind(ests, se)

```


These are the fitness and standard errors for HG and LG treatments (across densities)
```{r}
HG
LG
```

#Comparing Female (seeds set) and Male (seeds sired) Fitness

##Calculate mean seed set for each treat x N~e~ treatment to "relativize" female fitness

Calculate mean seed set for each individual plot. These values will be used to relativize fitness estimates
```{r}
aggregate(fin$seeds, by=list(fin$plotID), mean)

```

Load data with number of seeds sired (male fitness), where sires is the number of seeds sired for each family
```{r}
fin<- read.csv("data/aster.sire.dat.csv")

names(fin)
```

Note that the file `aster.sire.dat.csv` only has data from the High N~e~ treatment. We are only working with High N~e~ (2 full-sib individuals from 6 families), as we could not always asign paternity between two full-sibs. So, we assigned paternity to families and not individual plants.


Sum of number of seeds that we successfully assigned paternity to, per density treatment
```{r}
aggregate(fin$sires, by=list(fin$Den), sum)

```


Make sure class of factor variables for predictors (as was done with female fintess.
```{r}
fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)
```

The following generally follows the steps for performing aster analyses of female fitness (seeds set)

Load aster package
```{r}
library(aster)
```

Set the variables to be included in the graphical model. Notice that these are the same variables: flower number -> total fruit -> subsampled fruit -> seeds set
```{r}
vars<- c( "flw", "frt", "frt.2","seeds")

```


Subset data into High, Medium, and Low density treatments
```{r}
datH<- subset(fin, Den=="H")
datH$familyID<- droplevels(datH$familyID)
datM<- subset(fin, Den=="M")
datM$familyID<- droplevels(datM$familyID)
datL<- subset(fin, Den=="L")
datL$familyID<- droplevels(datL$familyID)
```

Perform the same data reshaping steps as before, reshaping data so that all response variables are located in a single vector in a new data sets called for each density treatment
```{r}
redataH <- reshape(datH, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
redataM <- reshape(datM, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
redataL <- reshape(datL, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
```

Designation of fitness variable, and add to each reshaped data.
```{r}
fit <- grepl("seeds", as.character(redataH$varb))
fit<- as.numeric(fit)

redataH$fit <- fit
redataM$fit <- fit
redataL$fit <- fit
```

Check organization of graphical models
```{r}
with(redataH, sort(unique(as.character(varb)[fit == 0])))
with(redataH, sort(unique(as.character(varb)[fit == 1])))

with(redataM, sort(unique(as.character(varb)[fit == 0])))
with(redataM, sort(unique(as.character(varb)[fit == 1])))

with(redataL, sort(unique(as.character(varb)[fit == 0])))
with(redataL, sort(unique(as.character(varb)[fit == 1])))
```



Setting the varible "root" to each redata file differs from the previous aster analyses. Here "root" is given the value 2 here to compliment the male fitness estimates (see section: Male Fitness).

```{r}
redataH<- data.frame(redataH, root=2)
redataM<- data.frame(redataM, root=2)
redataL<- data.frame(redataL, root=2)
```


Set graphical model and family for each node
```{r}
pred<- c(0,1,2,3)
fam<- c(1,2,2,2)
```

Perform aster analyses for three density treatments with `familyID` as fixed-effect
```{r}
aoutH<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataH)
summary(aoutH, show.graph = T)

aoutM<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataM)
summary(aoutM, show.graph = T)

aoutL<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataL)
summary(aoutL, show.graph = T)
```

##Generate Fitness Estimates for Each Genetic Family (familyID) in Each Density Treatment 

High Density Treatment Fitness Estimate

These steps follow those that produced mean fitness estimates for seed set 
```{r}
pout<- predict.aster(aoutH, se.fit=TRUE)

#make design matrix
fred <- data.frame(familyID=levels(redataH$familyID), flw=1, frt=1, frt.2=1, seeds=1,root = 2)

#reshape the design matrix just as the actual data
renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")

#make character string from "varb" of renewdata,
#without actual values (e.g., the layers of varb in renewdata)
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

#add layer to renewdata
renewdata<- data.frame(renewdata, layer= layer)

#add fit (seeds) in new layer col of renewdata as numeric, called fit
fit<- as.numeric(layer=="seeds")

#add fit to renewdata
renewdata<- data.frame(renewdata, fit = fit)

#rerun prediction of aout, with "made up" renewdata
pout<- predict(aoutH, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)

sapply(pout, class)

sapply(pout, length)

#put the parameter estimates into a matrix with individuals in rows
#and nodes along columns
nnode<- length(vars)
sally<- matrix(pout$fit, ncol = nnode)
dim(sally)# makes 10 x 4 matrix: 10 families by 4 nodes

#name the rows (by familyID) and columns (as nodes)
rownames(sally)<- unique(as.character(renewdata$familyID))
colnames(sally)<- unique(as.character(renewdata$varb))

#view matrix 
round(sally, 3)

#use just totalseeds as predicted (expected) fitneses
herman<- sally[,grepl("seeds", colnames(sally))]

#Generate Standard Errors for these estimates

nFam<- nrow(fred)
nnode<- length(vars)
amat<- array(0, c(nFam, nnode, nFam))
dim(amat)# makes an 10 x 4 x 10 matrix


foo<- grepl("seeds", vars)
for(k in 1:nFam)
  amat[k, foo, k]<- 1

#use aout object, with renewdata, and amat format
pout.amat<- predict(aoutH, newdata= renewdata, varvar= varb,
                    idvar= id, root = root, se.fit=TRUE, amat = amat)

#pout.amat$fit should be the same as file "herman"
herman
pout.amat$fit #they are the same.  Good.

#combine std.err with estimates, and then round
#to three decimal places
foo<- cbind(pout.amat$fit, pout.amat$se.fit)
rownames(foo)<- as.character(fred$familyID)
colnames(foo)<- c("High Den Fitness", "SE")
round(foo, 3) 
H_estimates<- round(foo,3)

H_estimates





```

Medium Density Fitness Estiamtes
```{r}
#Generate MLE of saturated model mean value parameter vector: mu
pout<- predict.aster(aoutM, se.fit=TRUE)


#make design matrix
fred <- data.frame(familyID=levels(redataM$familyID), flw=1, frt=1, frt.2=1, seeds=1,root = 2)

#reshape the "made up data" just as the actual data
renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")

#make character string from "varb" of renewdata,
#without actual values (e.g., the layers of varb in renewdata)
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

#add layer to renewdata
renewdata<- data.frame(renewdata, layer= layer)

#seed seed.ct in new layer col of renewdata as numeric, called fit
fit<- as.numeric(layer=="seeds")

#add fit to renewdata
renewdata<- data.frame(renewdata, fit = fit)

#rerun prediction of aout, with "made up" renewdata
pout<- predict(aoutM, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)

sapply(pout, class)
sapply(pout, length)

#put the parameter estimates into a matrix with familyID in rows
#and nodes along columns
nnode<- length(vars)
sally<- matrix(pout$fit, ncol = nnode)
dim(sally)# makes 9 x 4 matrix: 9 families by 4 nodes

#name the rows (by Den Treat) and columns (as nodes)
rownames(sally)<- unique(as.character(renewdata$familyID))
colnames(sally)<- unique(as.character(renewdata$varb))

#view matrix 
round(sally, 3)

#use just totalseeds as predicted (expected) fitneses
herman<- sally[,grepl("seeds", colnames(sally))]

#Generate Standard Errors for these estimates

nFam<- nrow(fred)
nnode<- length(vars)
amat<- array(0, c(nFam, nnode, nFam))
dim(amat)# makes an 9 x 4 x 9 matrix


foo<- grepl("seeds", vars)
for(k in 1:nFam)
  amat[k, foo, k]<- 1

#use aout object, with renewdata, and amat format
pout.amat<- predict(aoutM, newdata= renewdata, varvar= varb,
                    idvar= id, root = root, se.fit=TRUE, amat = amat)

#pout.amat$fit should be the same as file "herman"
herman
pout.amat$fit #they are the same.  Good.

#combine std.err with estimates, and then round
#to three decimal places
foo<- cbind(pout.amat$fit, pout.amat$se.fit)
rownames(foo)<- as.character(fred$familyID)
colnames(foo)<- c("Med Den Fitness", "SE")
round(foo, 3) 
M_estimates<- round(foo, 3)

M_estimates
```


Low Densitiy Estimates
```{r}

#generate MLE of saturated model mean value parameter vector: mu
pout<- predict.aster(aoutL, se.fit=TRUE)


#make data.frame of indivudals for each block (1-8)
fred <- data.frame(familyID=levels(redataL$familyID), flw=1, frt=1, frt.2=1, seeds=1,root = 2)

#reshape the "made up data" just as the actual data
renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")

#make character string from "varb" of renewdata,
#without actual values (e.g., the layers of varb in renewdata)
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

#add layer to renewdata
renewdata<- data.frame(renewdata, layer= layer)

#seed seed.ct in new layer col of renewdata as numeric, called fit
fit<- as.numeric(layer=="seeds")

#add fit to renewdata
renewdata<- data.frame(renewdata, fit = fit)

#rerun prediction of aout, with "made up" renewdata
pout<- predict(aoutM, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)

sapply(pout, class)
sapply(pout, length)

#put the parameter estimates into a matrix with familyID in rows
#and nodes along columns
nnode<- length(vars)
sally<- matrix(pout$fit, ncol = nnode)
dim(sally)# makes 8 x 4 matrix: 8 families by 4 nodes

#name the rows (by Den Treat) and columns (as nodes)
rownames(sally)<- unique(as.character(renewdata$familyID))
colnames(sally)<- unique(as.character(renewdata$varb))

#view matrix 
round(sally, 3)

#use just totalseeds as predicted (expected) fitneses
herman<- sally[,grepl("seeds", colnames(sally))]

#Generate Standard Errors for these estimates

nFam<- nrow(fred)
nnode<- length(vars)
amat<- array(0, c(nFam, nnode, nFam))
dim(amat)# makes an 8 x 4 x 8 matrix


foo<- grepl("seeds", vars)
for(k in 1:nFam)
  amat[k, foo, k]<- 1

#use aout object, with renewdata, and amat format
pout.amat<- predict(aoutL, newdata= renewdata, varvar= varb,
                    idvar= id, root = root, se.fit=TRUE, amat = amat)

#pout.amat$fit should be the same as file "herman"
herman
pout.amat$fit #they are the same.  Good.

#combine std.err with estimates, and then round
#to three decimal places
foo<- cbind(pout.amat$fit, pout.amat$se.fit)
rownames(foo)<- as.character(fred$familyID)
colnames(foo)<- c("Low Den Fitness", "SE")
round(foo, 3) 
L_estimates<- round(foo, 3)

L_estimates
```


The files present the mean fitness and standard error for each genetic family, in each density treatments under high N~e~ conditions
```{r}
H_estimates
M_estimates
L_estimates
```

# Estimate Male (Seeds Sired) Fitness

This section performs aster analyses with the number of seeds sired as the terminal fitness node, to obtain mean fitness estiames through male reproductive function. Because these analyses use the same data files as above (for relativized female fitness estimates), we can begin the process of setting up the aster analyses at designating the graphical model variables.

The graphical model for male fitness: flower number -> fruit number -> number of seed sired
```{r}
vars<- c( "flw", "frt", "sires")
```

Reshape data, as with female fitness, so that all response variables are located in a single vector in a new data
```{r}
redata <- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")
```


Designation of number of seeds sired ("sires") as fitness variable
```{r}
fit <- grepl("sires", as.character(redata$varb))
fit<- as.numeric(fit)

redata$fit <- fit

with(redata, sort(unique(as.character(varb)[fit == 0])))
with(redata, sort(unique(as.character(varb)[fit == 1])))
```



Add a variable "root" to redata, where value is 2

```{r}
redata<- data.frame(redata, root=2)
```


Set graphical model and statistical family for each node
```{r}
pred<- c(0,1,2)
fam<- c(1,2,2)

sapply(fam.default(), as.character)[fam]
```

## Main Aster Analysis of Male Fitness

First analysis with only fitness data, then add `familyID`, and finally density treatment. Perform liklihood ratio test to assess significance of `familyID` and `Den` 
```{r}
aout<- aster(resp~varb, pred, fam, varb, id, root, data=redata)

summary(aout)

aout2<- aster(resp~varb +fit:familyID, pred, fam, varb, id, root, data=redata)

summary(aout2)
```

Effect of `familyID` is not significant
```{r}
anova(aout, aout2)
```

Aster analsyis with `Den`
```{r}
aout3<- aster(resp~varb +fit:Den, pred, fam, varb, id, root, data=redata)


summary(aout3)
```

Effect of `Den` is not significant
```{r}
anova(aout, aout3)
```

##Family-specific Male Fitness Estimates

To compare with family-specific estimates of female fitness, we now generate family-specific male fitness estimates.

Load the data with male fitess (seeds sired), make sure all predictors are factors, and set the graphical model: flower number -> fruit number -> number of seed sired

```{r}
fin<- read.csv("C:data/aster.sire.dat.csv")

fin$Den<- as.factor(fin$Den)
fin$Gen<- as.factor(fin$Gen)
fin$plotID<- as.factor(fin$plotID)
fin$plantID<- as.factor(fin$plantID)
fin$familyID<- as.factor(fin$familyID)

vars<- c( "flw", "frt", "sires")
```




Reshape data so that all response variables are located in a single vector in a new data, and add `fit` and `root` to redata.
```{r}
redata <- reshape(fin, varying = list(vars), direction = "long",timevar = "varb", times = as.factor(vars), v.names = "resp")


fit <- grepl("sires", as.character(redata$varb))
fit<- as.numeric(fit)

redata$fit <- fit

with(redata, sort(unique(as.character(varb)[fit == 0])))
with(redata, sort(unique(as.character(varb)[fit == 1])))

redata<- data.frame(redata, root=2)
```


Set graph. model and family for each node
```{r}
pred<- c(0,1,2)
fam<- c(1,2,2)
```

Split reshaped data file into density-specific files for individual aster analsyses. Drop unused levels
```{r}
redataL<-subset(redata, Den=="L")

redataL$familyID<- droplevels(redataL$familyID)

redataM<- subset(redata, Den=="M")

redataM$familyID<- droplevels(redataM$familyID)

redataH<- subset(redata, Den=="H")

redataH$familyID<- droplevels(redataH$familyID)
```


### Low Density Fitness Estiamte for Male Fitness

Aster analysis for low density treatment
```{r}
aoutL<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataL)

summary(aoutL, show.graph = T)
```

Generate MLE of saturated model mean value parameter vector: mu
```{r}
pout<- predict.aster(aoutL, se.fit=TRUE)
```



Make a design matrix and reshape it just like the origional data
```{r}
fred <- data.frame(familyID=levels(redataL$familyID), flw=1, frt=1, sires=1,root = 2)

renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")
```



Ad `varb` and `fit` objects to reshaped design matrix
```{r}
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

renewdata<- data.frame(renewdata, layer= layer)

fit<- as.numeric(layer=="sires")

renewdata<- data.frame(renewdata, fit = fit)
```


Generate predicted values from aster analsyis (low density data_, with "made up" design matrix
```{r}
pout<- predict(aoutL, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)
```

Put the parameter estimates into object `ests` and standard errors into object `se`, that are both matrices with individuals in rows
and nodes along columns (makes 8 x 3 matrix: 8 indiv by 3 nodes).
```{r}
nnode<- length(vars)
ests<- matrix(pout$fit, ncol = nnode)
se<- matrix(pout$se.fit, ncol= nnode)
```

Name the rows (as familID) and columns (as nodes)
```{r}
rownames(ests)<- unique(as.character(renewdata$familyID))
colnames(ests)<- unique(as.character(renewdata$varb))

rownames(se)<- unique(as.character(renewdata$familyID))
colnames(se)<- unique(as.character(renewdata$varb))

round(ests, 3)

round(se, 3)
```

Use just seeds sired as predicted fitneses estiamte
```{r}
L_ests<- ests[,grepl("sires", colnames(ests))]
L_se<- se[,grepl("sires", colnames(se))]
```

Combine estimates and standard errors into single object
```{r}
L_fit<- cbind(L_ests, L_se)

round(L_fit, 3)
```


### Medium Density Fitness Estiamte for Male Fitness

Aster analysis for medium density treatment
```{r}
aoutM<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataM)

summary(aoutM, show.graph = T)
```

Generate MLE of saturated model mean value parameter vector: mu
```{r}
pout<- predict.aster(aoutM, se.fit=TRUE)
```



Make a design matrix and reshape it just like the origional data
```{r}
fred <- data.frame(familyID=levels(redataL$familyID), flw=1, frt=1, sires=1,root = 2)

renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")
```



Ad `varb` and `fit` objects to reshaped design matrix
```{r}
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

renewdata<- data.frame(renewdata, layer= layer)

fit<- as.numeric(layer=="sires")

renewdata<- data.frame(renewdata, fit = fit)
```


Generate predicted values from aster analsyis (low density data_, with "made up" design matrix
```{r}
pout<- predict(aoutM, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)
```

Put the parameter estimates into object `ests` and standard errors into object `se`, that are both matrices with individuals in rows
and nodes along columns (makes 8 x 3 matrix: 8 indiv by 3 nodes).
```{r}
nnode<- length(vars)
ests<- matrix(pout$fit, ncol = nnode)
se<- matrix(pout$se.fit, ncol= nnode)
```

Name the rows (as familID) and columns (as nodes)
```{r}
rownames(ests)<- unique(as.character(renewdata$familyID))
colnames(ests)<- unique(as.character(renewdata$varb))

rownames(se)<- unique(as.character(renewdata$familyID))
colnames(se)<- unique(as.character(renewdata$varb))

round(ests, 3)

round(se, 3)
```

Use just seeds sired as predicted fitneses estiamte
```{r}
M_ests<- ests[,grepl("sires", colnames(ests))]
M_se<- se[,grepl("sires", colnames(se))]
```

Combine estimates and standard errors into single object
```{r}
M_fit<- cbind(M_ests, M_se)

round(M_fit, 3)
```

### High Density Fitness Estiamte for Male Fitness

Aster analysis for high density treatment
```{r}
aoutH<- aster(resp~varb + fit:(familyID), pred, fam, varb, id, root, data=redataH)

summary(aoutH, show.graph = T)
```

Generate MLE of saturated model mean value parameter vector: mu
```{r}
pout<- predict.aster(aoutH, se.fit=TRUE)
```



Make a design matrix and reshape it just like the origional data
```{r}
fred <- data.frame(familyID=levels(redataL$familyID), flw=1, frt=1, sires=1,root = 2)

renewdata <- reshape(fred, varying = list(vars),
                     direction = "long", timevar = "varb",
                     times = as.factor(vars), v.names = "resp")
```



Ad `varb` and `fit` objects to reshaped design matrix
```{r}
layer<- gsub("[0-9]", "", as.character(renewdata$varb))

renewdata<- data.frame(renewdata, layer= layer)

fit<- as.numeric(layer=="sires")

renewdata<- data.frame(renewdata, fit = fit)
```


Generate predicted values from aster analsyis (low density data_, with "made up" design matrix
```{r}
pout<- predict(aoutH, newdata= renewdata, varvar= varb,
               idvar = id, root = root, se.fit = TRUE)
```

Put the parameter estimates into object `ests` and standard errors into object `se`, that are both matrices with individuals in rows
and nodes along columns (makes 8 x 3 matrix: 8 indiv by 3 nodes).
```{r}
nnode<- length(vars)
ests<- matrix(pout$fit, ncol = nnode)
se<- matrix(pout$se.fit, ncol= nnode)
```

Name the rows (as familID) and columns (as nodes)
```{r}
rownames(ests)<- unique(as.character(renewdata$familyID))
colnames(ests)<- unique(as.character(renewdata$varb))

rownames(se)<- unique(as.character(renewdata$familyID))
colnames(se)<- unique(as.character(renewdata$varb))

round(ests, 3)

round(se, 3)
```

Use just seeds sired as predicted fitneses estiamte
```{r}
H_ests<- ests[,grepl("sires", colnames(ests))]
H_se<- se[,grepl("sires", colnames(se))]
```

Combine estimates and standard errors into single object
```{r}
H_fit<- cbind(H_ests, H_se)

round(H_fit, 3)
```

# Above and Below Ground Biomass 

## Preliminary and Summary of Data

The following code summarizes and performs linear analyses of biomass accumulation (above and below ground). The required data is in the same file as the female fitness data.

Reload data and load package `emmeans`
```{r}
fin<- read.csv("C:data/aster.dat.csv")

library(emmeans)


```
Subset the data into High and Low N~e~ data
```{r}
hi<- subset(fin, Gen=="HG")
lo<- subset(fin, Gen=="LG")

```

For summary purposes, define a function that will calculate standard errors
```{r}
stderr<- function(x) sd(x)/sqrt(length(x))
```

###Generate Summary Statistics (mean & standard error) for above and below biomass, in High and Low N~e~ plants

High N~e~

Above ground biomass
```{r}
aggregate(hi$mass.a, by=list(hi$Den), mean)
aggregate(hi$mass.a, by=list(hi$Den), stderr)
```

Below ground biomass
```{r}
aggregate(hi$mass.b, by=list(hi$Den), mean)
aggregate(hi$mass.b, by=list(hi$Den), stderr)
```

Low N~e~
Above ground biomass
```{r}
aggregate(lo$mass.a, by=list(lo$Den), mean)
aggregate(lo$mass.a, by=list(lo$Den), stderr)
```

Below ground biomass
```{r}
aggregate(lo$mass.b, by=list(lo$Den), mean)
aggregate(lo$mass.b, by=list(lo$Den), stderr)
```

##Linear Analysis of Biomass
###Above Ground Biomass (`mass.a`)

Sequentially add Density, N~e~, and Den x N~e~ interaction. Note all analyses log-transform biomass.
```{r}
f.lm<- lm(log(mass.a) ~ (Den) , data=fin)

f.lm2<- lm(log(mass.a) ~ (Den + Gen) , data=fin)

f.lm3<- lm(log(mass.a) ~ (Den + Gen + Den*Gen) , data=fin)

summary(f.lm)
summary(f.lm2)
summary(f.lm3)
```

Perform liklihood ratio test of models with Density, N~e~, and Den x N~e~ interaction
```{r}
anova(f.lm, f.lm2, f.lm3)
```
Generate Least Square Means (EMeans) for Density treatmetns across High and Low N~e~

```{r}

emmeans(f.lm2, "Den", type='response', by="Gen")

pairs(emmeans(f.lm2, "Den", type='response', by="Gen"))
test(emmeans(f.lm2, "Den", type='response', by="Gen"))
```


###Below Ground Biomass (`mass.b`)
Sequentially add Density, N~e~, and Den x N~e~ interaction. Note all analyses log-transform biomass.
```{r}
b.lm<- lm(log(mass.b) ~ (Den) , data=fin)

b.lm2<- lm(log(mass.b) ~ (Den + Gen) , data=fin)

b.lm3<- lm(log(mass.b) ~ (Den + Gen + Den*Gen) , data=fin)

summary(b.lm)
summary(b.lm2)
summary(b.lm3)
```


Perform liklihood ratio test of models with Density, N~e~, and Den x N~e~ interaction.
```{r}
anova(b.lm, b.lm2, b.lm3)
```

Generate Least Square Means (EMeans) for Density treatmetns across High and Low N~e~
```{r}
emmeans(b.lm2, "Den", type='response', by="Gen")

pairs(emmeans(b.lm2, "Den", type='response', by="Gen"))
test(emmeans(b.lm2, "Den", type='response', by="Gen"))
```


# Aborted Ovules

## Preliminary and Summary of Data

The following code summarizes and performs linear analyses of the number of aborted ovules. The required data is in the same file as the female fitness data.

Reload data and load package `emmeans`
```{r}
fin<- read.csv("C:data/aster.dat.csv")

library(emmeans)

```

Perform linear analysis of aborted ovules (log transformed) with Density, Density + N~e~, and Density + N~e~ + Density x N~e~ in sequential models. 
```{r}
lm1<- lm(log(aborted +1) ~ Den , data=fin)

lm2<- lm(log(aborted +1) ~ Den + Gen, data=fin)

lm3<- lm(log(aborted +1) ~ Gen + Den + Den*Gen, data=fin)

summary(lm1)
summary(lm2)
summary(lm3)
```

Perform liklihood ratio test of above models
```{r}
anova(lm1, lm2, lm3)

```

Finally, generate least square means 
```{r}
emmeans(lm3, "Den", "Gen", type="response")
pairs(emmeans(lm3, "Den","Gen", type="response"))
test(emmeans(lm3, "Den","Gen", type="response"))

```

